---
title: "rstan_tuto"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = NA, 
  cache = FALSE,
  fig.width = 12,
  fig.height = 8,
  message = FALSE,
  dev = c("png", "pdf")
)
options(width = 100)

library("rstan")
library("mcmc")
```

# A first basic example

```{r, warning=FALSE}
writeLines(readLines("mvn.stan"))
```

```{r, warning=FALSE}
fit_mvn <- stan(file = "mvn.stan",
                data = list(d=8), algorithm = "HMC",
                iter = 400, warmup = 200, 
                chains = 4, refresh = 0)
fit_mvn
```

```{r, warning=FALSE}
stan_plot(fit_mvn)
stan_trace(fit_mvn)
stan_hist(fit_mvn)
stan_dens(fit_mvn)
stan_diag(fit_mvn)
stan_rhat(fit_mvn)
stan_ess(fit_mvn)
```


# Bayesian logistic regression

\begin{equation*}
    \boldsymbol{\beta} \sim \mathcal{N}(0, 0.35^2. \boldsymbol{I}_4), \quad
    y_j \sim \text{Bernoulli}\left(\frac{1}{1+e^{-\boldsymbol{x}_j^\top\boldsymbol{\beta}}}\right).
\end{equation*}


```{r, warning=FALSE}
writeLines(readLines("logit_reg.stan"))
```

We run Stan using the dataset \texttt{logit} available in the package \texttt{mcmc}.


```{r, warning=FALSE, message=FALSE, override = FALSE}
data("logit")

head(logit)
```


```{r, warning=FALSE, message=FALSE, override = FALSE}
logit_data<- list(D = 4,
                  N = 100,
                  tau = 0.35,
                  x = logit[c("x1", "x2", "x3", "x4")],
                  y = logit[["y"]])

fit_logit <- stan(file = "logit_reg.stan",
                  data=logit_data, algorithm = "HMC",
                  iter=400, chains=4, refresh=0)
fit_logit
```
```{r, warning=FALSE}
stan_plot(fit_logit)
stan_trace(fit_logit)
stan_hist(fit_logit)
stan_dens(fit_logit)
stan_diag(fit_logit)
stan_rhat(fit_logit)
stan_ess(fit_logit)
```

# Hierarchical model: Eight Schools


```{r}
eight_schools <- list(J = 8,
                      y = c(28,  8, -3,  7, -1,  1, 18, 12),
                      sigma = c(15, 10, 16, 11,  9, 11, 10, 18))
```

\begin{tabular}{ll}
    \textbf{Example~\ref{fig:hierarchical}.a:} & Centered parameterisation (CP) \\[0.1cm]
    & $\theta_j \sim \mathcal{N}(\mu,\tau), \quad 
    y_j \sim \mathcal{N}(\theta_j,\sigma^2_j)$.\\[0.2cm]
    \textbf{Example~\ref{fig:hierarchical}.b:} & Non-centered parameterisation (NCP) \\[0.1cm]
    & $\bar{\theta}_j \sim \mathcal{N}(0,1), \quad 
    \theta_j = \mu + \tau \bar{\theta}_j, \quad y_j \sim \mathcal{N}(\theta_j,\sigma^2_j)$.\\[0.2cm]
\end{tabular}


## Centered model


```{r, warning=FALSE}
writeLines(readLines("eight_schools_cp.stan"))
```


```{r, warning=FALSE}
fit_cp <- stan(
  file = "eight_schools_cp.stan", data = eight_schools,
  iter = 1000, chains = 4, refresh = 0
)

fit_cp
```

```{r, warning=FALSE}
stan_plot(fit_cp)
stan_trace(fit_cp)
stan_hist(fit_cp)
stan_dens(fit_cp)
stan_diag(fit_cp)
stan_rhat(fit_cp)
stan_ess(fit_cp)
```

### Non-centered model

```{r}
writeLines(readLines("eight_schools_ncp.stan"))
```

```{r, warning=FALSE}
fit_ncp <- stan(
  file = "eight_schools_ncp.stan", data = eight_schools,
  iter = 1000, chains = 4, refresh = 0
)

fit_ncp
```


```{r, warning=FALSE}
stan_plot(fit_ncp)
stan_trace(fit_ncp)
stan_hist(fit_ncp)
stan_dens(fit_ncp)
stan_diag(fit_ncp)
stan_rhat(fit_ncp)
stan_ess(fit_ncp)
```
